Low Infant Birth Weight in Brazil

### 2024-06-24 code/analysis-code/fullanalysis-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

#rm(list=ls()) #clear the environment
```

@@

# Save regression output into table
saveRDS(table2_train, file = here::here("results", "tables", "table2_train.rds"))
```
Our results here are not that different from the model run on all the data. Our adjusted R-squared is 0.007, the same as the full model. The RMSE is 495, which is slightly lower than the full model (502), which shows better performance.

@@

# Save regression output into table
saveRDS(ML_cv1_metrics, file = here::here("results", "tables", "ML_cv1_metrics.rds"))
```
Performing cross-validation results in an RMSE that is the same as our training model (495), lower than our full model. The r-squared (0.02) is also higher than our training model (0.008) and our full model (0.007), demonstrating better model performance.

@@

When running the model on our testing dataset, our results are not significant, given the small sample size.

@@

# Save regression output into table
saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
```
Our RMSE is 498 and our r-squared is 0.02.

@@

# Save regression output into table
saveRDS(table3_train, file = here::here("results", "tables", "table3_train.rds"))
```
Our RMSE is 490 and our r-squared is 0.03, a slightly better performance than the full model.

@@

# Save regression output into table
saveRDS(ML_cv2_metrics, file = here::here("results", "tables", "ML_cv2_metrics.rds"))
```
Here, our RMSE is 458, an improvement from both the full model and the training model, and our r-squared is 0.03, the same as our training model.

@@

# Save regression output into table
saveRDS(table3_test, file = here::here("results", "tables", "table3_test.rds"))
```
Running the model on our test data, we see that the only significant result is for the nulliparous status. All other results are insignificant.

@@

Our RMSE is 502 and our r-squared is 0.007.

@@

Our RMSE is 495 and our r-squared is 0.009, slightly better performance than the full model. Our results are not significantly different from the full model.

@@

# Save regression output into table
saveRDS(ML_cv3_metrics, file = here::here("results", "tables", "ML_cv3_metrics.rds"))
```
Using cross-validation, our RMSE is 495 and our r-squared is 0.02, a similar performance to our training model.

@@

Much like in the other testing models, our results are not significant (except the intercept). Our performance is also worse (RMSE of 518 and r-squared of 0.005).

@@

# Save regression output into table
saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
```
Our RMSE is 498 and our r-squared is 0.02.

@@

# Save regression output into table
saveRDS(table5_train, file = here("results", "tables", "table5_train.rds"))
```
Our RMSE is 490 and our r-squared is 0.03, a slightly better performance than the full model.

@@

# Save regression output into table
saveRDS(ML_cv4_metrics, file = here::here("results", "tables", "ML_cv4_metrics.rds"))
```
Here, our RMSE is 458 and our r-squared is 0.03, a better performance than both the full model and training model.

@@

# Save regression output into table
saveRDS(table5_test, file = here("results", "tables", "table5_test.rds"))
```
Again, most of our results are not significant except the intercept and nulliparous variable given the small sample size.

>>>>

rm(list=ls()) #clear the environment
```

@@

# Save regression output into table
saveRDS(table2_train, file = here::here("results", "tables", "table2_train.rds"))
```
Our results here are not that different from the model run on all the data. Our adjusted R-squared is 0.01, the same as the full model. The RMSE is 488, which is slightly lower than the full model (494), which shows better performance.

@@

# Save regression output into table
saveRDS(ML_cv1_metrics, file = here::here("results", "tables", "ML_cv1_metrics.rds"))
```
Performing cross-validation results in an RMSE that is the same as our training model (488), lower than our full model. The r-squared (0.015) is also higher than our training and full models (0.01), demonstrating better model performance.

@@

When running the model on our testing dataset, our results remain significant at the 0.05 level. The RMSE is 508 and the r-squared is 0.008. This shows a slightly worse fit than our training results.

@@

# Save regression output into table
saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
```
Our RMSE is 490 and our r-squared is 0.026.

@@

# Save regression output into table
saveRDS(table3_train, file = here::here("results", "tables", "table3_train.rds"))
```
Our RMSE is 483 and our r-squared is 0.03, a slightly better performance than the full model.

@@

# Save regression output into table
saveRDS(ML_cv2_metrics, file = here::here("results", "tables", "ML_cv2_metrics.rds"))
```
Here, our RMSE is 483 and our r-squared is 0.026, the same as the training model.

@@

# Save regression output into table
saveRDS(table3_test, file = here::here("results", "tables", "table3_test.rds"))
```
Running the model on our test data, we see no significant results at any alpha level.

@@

Our RMSE is 494 and our r-squared is 0.01.

@@

Our RMSE is 489 and our r-squared is 0.01, slightly better performance than the full model. Our results are not significantly different from the full model.

@@

# Save regression output into table
saveRDS(ML_cv3_metrics, file = here::here("results", "tables", "ML_cv3_metrics.rds"))
```
Using cross-validation, our RMSE is 487 and our r-squared is 0.01, a similar performance to our training model.

@@

Much like in the other testing models, our results are not significant (except the intercept). Our performance is also worse (RMSE of 507 and r-squared of 0.008).

@@

# Save regression output into table
saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
```
Our RMSE is 490 and our r-squared is 0.026.

@@

# Save regression output into table
saveRDS(table5_train, file = here("results", "tables", "table5_train.rds"))
```
Our RMSE is 483 and our r-squared is 0.033, a slightly better performance than the full model.

@@

# Save regression output into table
saveRDS(ML_cv4_metrics, file = here::here("results", "tables", "ML_cv4_metrics.rds"))
```
Here, our RMSE is 484 and our r-squared is 0.025, a better performance than both the full model and training model.

@@

# Save regression output into table
saveRDS(table5_test, file = here("results", "tables", "table5_test.rds"))
```
Again, most of our results are not significant except the intercept given the small sample size.

### 2024-06-24 code/analysis-code/introanalysis-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling

#rm(list=ls()) #clear the environment
```

@@

@@

For all three models, the r-squared is very small (0.0073, 0.0061, and 0.027), indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

@@

>>>>

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling
rm(list=ls()) #clear the environment
```

@@

## Linear Models

@@

For all three models, the r-squared is very small (0.009, 0.008, and 0.026), indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

@@

## Logistic models

Now we will still consider `Birthweight` as the outcome of interest, but we will transform it into a categorical outcome of <=2500 (low birthweight, LBW) and >2500 (normal birthweight, NBW) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.

First, we will split data into training and testing sets to later calculate ROC_AUC.

```{r}
set.seed(123)
ML_log_split <- initial_split(ML_linear, prop = 0.75, strata = BirthweightCategory)

# Create training and test datasets
ML_log_train <- training(ML_log_split)
ML_log_test <- testing(ML_log_split)
```

Now we will run the first logistic model with `ModifiedColor` as the predictor.

```{r}
# Simple logistic regression model with ModifiedColor as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW

# Set the reference level explicitly to NBW
ML_linear <- ML_linear %>%
  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

# Define model specification
log1_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Create workflow
log1_wf <- workflow() %>%
  add_recipe(log1_recipe) %>%
  add_model(log1_spec)

# Fit model
log1_fit <- log1_wf %>%
fit(data = ML_linear)

# Extract fitted model from workflow
log1_fitted <- extract_fit_parsnip(log1_fit)$fit

# Create a gtsummary table directly from the model
table6 <- tbl_regression(log1_fitted, exponentiate = TRUE) # use exponentiate true for ORs
table6

# Save regression output into table
saveRDS(table6, file = here::here("results", "tables", "table6.rds"))

# Calculate performance metrics using pseudo-R2
# Extract model metrics using broom
model_metrics1 <- glance(log1_fitted)

# Calculate McFadden's R-squared
# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
mcfadden_r2_1 <- 1 - (model_metrics1$deviance / model_metrics1$null.deviance)

# Print McFadden's R-squared
print(mcfadden_r2_1)

# Calculate performance metrics using ROC_AUC
# Fit the model on the training data
log1_fit <- log1_wf %>% fit(data = ML_log_train)

# Predict probabilities on the test set
prob_preds1 <- predict(log1_fit, new_data = ML_log_test, type = "prob")

# Combine predictions with true values
results1 <- bind_cols(ML_log_test, prob_preds1)

# Calculate ROC-AUC
roc_auc_result1 <- roc_auc(results1, truth = BirthweightCategory, .pred_LBW)

# Print ROC-AUC
print(roc_auc_result1)

# Create and plot the ROC curve
roc_curve_data1 <- roc_curve(results1, truth = BirthweightCategory, .pred_LBW)

# Plotting the ROC curve
roc_plot1 <- ggplot(roc_curve_data1, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 2) +
  labs(title = "ROC Curve",
       x = "1 - Specificity",
       y = "Sensitivity")

# Print the plot
print(roc_plot1)
```
With an ROC_AUC of 0.463, our model is performing worse than random guessing. Our pseudo R-squared is 0.004.

Now let's run another logistic model with `Color` as the predictor.

```{r}
# Simple logistic regression model with ModifiedColor as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW

# Set the reference level explicitly to NBW
ML_linear <- ML_linear %>%
  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log2_recipe <- recipe(BirthweightCategory ~ Color, data = ML_linear) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

# Define model specification
log2_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Create workflow
log2_wf <- workflow() %>%
  add_recipe(log2_recipe) %>%
  add_model(log2_spec)

# Fit model
log2_fit <- log2_wf %>%
fit(data = ML_linear)

# Extract fitted model from workflow
log2_fitted <- extract_fit_parsnip(log2_fit)$fit

# Create a gtsummary table directly from the model
table7 <- tbl_regression(log2_fitted, exponentiate = TRUE) # use exponentiate true for ORs
table7

# Save regression output into table
saveRDS(table7, file = here::here("results", "tables", "table7.rds"))

# Calculate metrics using pseudo-R2
# Extract model metrics using broom
model_metrics2 <- glance(log2_fitted)

# Calculate McFadden's R-squared
# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
mcfadden_r2_2 <- 1 - (model_metrics2$deviance / model_metrics2$null.deviance)

# Print McFadden's R-squared
print(mcfadden_r2_2)

# Calculate performance metrics using ROC_AUC
# Fit the model on the training data
log2_fit <- log2_wf %>% fit(data = ML_log_train)

# Predict probabilities on the test set
prob_preds2 <- predict(log2_fit, new_data = ML_log_test, type = "prob")

# Combine predictions with true values
results2 <- bind_cols(ML_log_test, prob_preds2)

# Calculate ROC-AUC
roc_auc_result2 <- roc_auc(results2, truth = BirthweightCategory, .pred_LBW)

# Print ROC-AUC
print(roc_auc_result2)

# Create and plot the ROC curve
roc_curve_data2 <- roc_curve(results2, truth = BirthweightCategory, .pred_LBW)

# Plotting the ROC curve
roc_plot2 <- ggplot(roc_curve_data2, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 2) +
  labs(title = "ROC Curve",
       x = "1 - Specificity",
       y = "Sensitivity")

# Print the plot
print(roc_plot2)
```
The ROC_AUC for our second simple logistic regression model is 0.47, or still worse than random guessing. Our pseudo R-squared is 0.006.

Now, we will run two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.

```{r}
# Multiple logistic regression model with ModifiedColor as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW
levels(ML_linear$ModifiedColor) # Reference is Euro-descent
levels(ML_linear$ModifiedStatus) # Reference is multiparous
levels(ML_linear$ModifiedNationality) # Reference is European

# Set the reference level explicitly to NBW
ML_linear <- ML_linear %>%
  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"),
         ModifiedNationality = relevel(ModifiedNationality, ref = "Brazilian"))

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log3_recipe <- recipe(BirthweightCategory ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

# Define model specification
log3_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Create workflow
log3_wf <- workflow() %>%
  add_recipe(log3_recipe) %>%
  add_model(log3_spec)

# Fit model
log3_fit <- log3_wf %>%
fit(data = ML_linear)

# Extract fitted model from workflow
log3_fitted <- extract_fit_parsnip(log3_fit)$fit

# Create a gtsummary table directly from the model
table8 <- tbl_regression(log3_fitted, exponentiate = TRUE) # use exponentiate true for ORs
table8 <- bold_p(table8, t = 0.05)
table8

# Save regression output into table
saveRDS(table8, file = here::here("results", "tables", "table8.rds"))

# Calculate metrics using pseudo-R2
# Extract model metrics using broom
model_metrics3 <- glance(log3_fitted)

# Calculate McFadden's R-squared
# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
mcfadden_r2_3 <- 1 - (model_metrics3$deviance / model_metrics3$null.deviance)

# Print McFadden's R-squared
print(mcfadden_r2_3)

# Calculate performance metrics using ROC_AUC
# Fit the model on the training data
log3_fit <- log3_wf %>% fit(data = ML_log_train)

# Predict probabilities on the test set
prob_preds3 <- predict(log3_fit, new_data = ML_log_test, type = "prob")

# Combine predictions with true values
results3 <- bind_cols(ML_log_test, prob_preds3)

# Calculate ROC-AUC
roc_auc_result3 <- roc_auc(results3, truth = BirthweightCategory, .pred_LBW)

# Print ROC-AUC
print(roc_auc_result3)

# Create and plot the ROC curve
roc_curve_data3 <- roc_curve(results3, truth = BirthweightCategory, .pred_LBW)

# Plotting the ROC curve
roc_plot3 <- ggplot(roc_curve_data3, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 2) +
  labs(title = "ROC Curve",
       x = "1 - Specificity",
       y = "Sensitivity")

# Print the plot
print(roc_plot3)
```
Our ROC_AUC here is worse, at 0.46, or worse than random guessing. Our pseudo-R2 is 0.0055.

```{r}
# Multiple logistic regression model with Color as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW
levels(ML_linear$Color) # Reference is White
levels(ML_linear$ModifiedStatus) # Reference is multiparous
levels(ML_linear$ModifiedNationality) # Reference is European

# Set the reference level explicitly to NBW
ML_linear <- ML_linear %>%
  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"),
         ModifiedNationality = relevel(ModifiedNationality, ref = "Brazilian"))

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log4_recipe <- recipe(BirthweightCategory ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

# Define model specification
log4_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Create workflow
log4_wf <- workflow() %>%
  add_recipe(log4_recipe) %>%
  add_model(log4_spec)

# Fit model
log4_fit <- log4_wf %>%
fit(data = ML_linear)

# Extract fitted model from workflow
log4_fitted <- extract_fit_parsnip(log4_fit)$fit

# Create a gtsummary table directly from the model
table9 <- tbl_regression(log4_fitted, exponentiate = TRUE) # exponentiate true for ORs

#Bold significant p-values
table9 <- bold_p(table9, t = 0.05)
table9

# Save regression output into table
saveRDS(table9, file = here::here("results", "tables", "table9.rds"))

# Calculate metrics using pseudo-R2
# Extract model metrics using broom
model_metrics4 <- glance(log4_fitted)

# Calculate McFadden's R-squared
# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
mcfadden_r2_4 <- 1 - (model_metrics4$deviance / model_metrics4$null.deviance)

# Print McFadden's R-squared
print(mcfadden_r2_4)

# Calculate performance metrics using ROC_AUC
# Fit the model on the training data
log4_fit <- log4_wf %>% fit(data = ML_log_train)

# Predict probabilities on the test set
prob_preds4 <- predict(log4_fit, new_data = ML_log_test, type = "prob")

# Combine predictions with true values
results4 <- bind_cols(ML_log_test, prob_preds4)

# Calculate ROC-AUC
roc_auc_result4 <- roc_auc(results4, truth = BirthweightCategory, .pred_LBW)

# Print ROC-AUC
print(roc_auc_result4)

# Create and plot the ROC curve
roc_curve_data4 <- roc_curve(results4, truth = BirthweightCategory, .pred_LBW)

# Plotting the ROC curve
roc_plot4 <- ggplot(roc_curve_data4, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 2) +
  labs(title = "ROC Curve",
       x = "1 - Specificity",
       y = "Sensitivity")

# Print the plot
print(roc_plot4)
```
In our final model, our ROC_AUC is 0.477, still worse than random guessing. Our pseudo-R2 is 0.007.

### 2024-06-24 code/eda-code/eda-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

@@

#Save table
saveRDS(table1, file = here::here("results", "tables", "table1_final.rds"))
```

>>>>

In the second part of this code chunk, I also break down infant birth weight by sex and by race. I will need to come back and rename these tables.

@@

#Save table
saveRDS(table1, file = here::here("results", "tables", "table1_final.rds"))

#Create gtsummary table on infant birth weight stratified by sex and maternal skin color
stratified_color <- 
  ML_summary %>%
  filter(!is.na(Weightgrams) & !is.na(ModifiedColor)) %>% # Exclude rows with NA in weight & MC variable
  rename("Infant Weight (grms)" = Weightgrams) %>% # Rename Weightgrams to Infant Weight (grms)
  tbl_summary(
    by = ModifiedColor,  # Grouping by ModifiedColor
    include = "Infant Weight (grms)",
    statistic = list("Infant Weight (grms)" ~ "{mean} ({sd})")
  )
stratified_color

stratified_sex <- 
  ML_summary %>%
  filter(!is.na(Weightgrams) & !is.na(Sex)) %>%
  rename("Infant Weight (grms)" = Weightgrams) %>% # Rename Weightgrams to Infant Weight (grms)
  tbl_summary(
    by = Sex,  # Grouping by Sex
    include = "Infant Weight (grms)",
    statistic = list("Infant Weight (grms)" ~ "{mean} ({sd})")
  )
stratified_sex

# Combine the two tables
table1.1 <- tbl_merge(
  tbls = list(stratified_color, stratified_sex),
  tab_spanner = c("**Weight by Modified Color**", "**Weight by Sex**"))

# Display the modified table
table1.1

#Save table
saveRDS(table1.1, file = here::here("results", "tables", "table1.1_final.rds"))
```

### 2024-06-24 code/processing-code/processingfile-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

```{r}
library(readxl) #for loading Excel files
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
```

@@

I created this dataset from published clinical notes from a specific maternity hospital in Rio de Janeiro, Brazil for multiple projects, mostly of a qualitative nature. Thus, I was interested in the notes clinicians included on difficult deliveries and their outcomes. These variables include qualitative information that can be excluded from our processed data, since we are not looking at qualitative notes on cases. We also don't need the citation information to run regressions. So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Birth_Notes`, `Maternal_Notes`, `Fetal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.

```{r}
# Remove unnecessary variables
ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Birth_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
```

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
```

We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations. This is a lot. Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.

```{r}
# Remove variables with missing data not relevant to hypothesis
ML_2 <- ML_1 %>% dplyr::select(-CivilStatus, -GestationalAge_Months)
```

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_2 %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(missing_data_summary)
```

@@

```{r}
# Transform character variables into factors
ML_3 <- ML_2 %>%
  mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex), as.factor))

# Check variables
glimpse(ML_3)

# Check levels of factor variables
map(ML_3[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')], levels)

# Define factor variables to check frequencies
factor_variables <- c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')

# Check frequencies of factor variables
frequency_table <- ML_3 %>%
  count(across(all_of(factor_variables)))

@@

There are still missing observations for all the remaining variables except `Date`. For now, I will keep these observations intact because I want the maximum number of observations for summary statistics of all variables. Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study. Here, I will create two datasets: one for summary statistics and one for the linear regression model. In addition, most contemporary studies exclude all birthweights less than 500 grams because they have a high chance of being stillborn. I do so the same here, even if they were marked as a live birth. I will also exclude all stillbirths or neonatal deaths as the analysis is interested in studying live births.

```{r} 
# Create dataset for summary statistics (includes missing values)
ML_summary <- ML_3

# Creating dataset for linear regression
ML_linear <- ML_3 %>% tidyr::drop_na() %>%
  filter(Weightgrams >= 500, FetalOutcome != "morto") # Drop observations of < 500 grams and all non-live births

# Check for no missing data in linear model
ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(ML_linear_summary)
```

Now, I can see that my `ML_linear` dataset has no missing data. It has 1944 observations and 11 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.

@@

Finally, I will save the clean data both the `ML_Linear` dataset and the `ML_summary` dataset as RDS files
.
```{r}
# Save ML_Linear as RDS
save_data_location <- here::here("data","processed-data","ML_linear_processed.rds")
saveRDS(ML_linear, file = save_data_location)

>>>>

```{r}
library(readxl) #for loading Excel files
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
rm(list=ls()) #clear the environment
```

@@

I created this dataset from published clinical notes from a specific maternity hospital in Rio de Janeiro, Brazil for multiple projects, mostly of a qualitative nature. Thus, I was interested in the notes clinicians included on difficult deliveries and their outcomes. These variables include qualitative information that can be excluded from our processed data, since we are not looking at qualitative notes on cases. We also don't need the citation information to run regressions. So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Maternal_Notes`, `Fetal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`. Further, we will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).

```{r}
# Remove unnecessary variables
ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
```

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```

We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations. This is a lot. Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth. `Birth_Notes` also has 2330 missing values, but first I need to extract information on gestation (twin versus singleton birth) from this variable before I exclude it.

```{r}
# Extract information on gestation from Birth_Notes and create new variable gestation with singleton and twin births
ML_2 <- ML_1 %>%
  mutate(Birth_Notes = str_trim(Birth_Notes)) %>% # removes any leading or trailing whitespace from Birth_Notes column
  mutate(Gestation = if_else(is.na(Birth_Notes) | Birth_Notes == "", "singleton", # checks if Birth_Notes is NA or empty strings and assigns singleton to Gestation column
                             if_else(str_detect(tolower(Birth_Notes), "gemelar"), "twin", "singleton"))) # checks for gemelar in case-insensitive manner and assigns twin accordingly or singleton otherwise
ML_2

# Remove variables with missing data not relevant to hypothesis
ML_3 <- ML_2 %>% dplyr::select(-CivilStatus, -GestationalAge_Months, -Birth_Notes)
```

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_3 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```
Now, I will create a new categorical variable of Normal Birthweight (NBW) for all birth weights greater than 2500 grams and Low Birthweight (LBW) for all birth weights equal to or less than 2500 grams for a future logistic regression. This will come from the `Weightgrams` variable.

```{r}
# Create new variable of NBW and LBW for future logistic regression
ML_4 <- ML_3 %>%
  mutate(BirthweightCategory = if_else(Weightgrams <= 2500, "LBW", "NBW"))
ML_4
```

@@

```{r}
# Transform character variables into factors
ML_5 <- ML_4 %>%
  mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex, Gestation, BirthweightCategory), as.factor))

# Check variables
glimpse(ML_5)

# Check levels of factor variables
map(ML_5[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex', 'Gestation', 'BirthweightCategory')], levels)

# Define factor variables to check frequencies
factor_variables <- c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex', 'Gestation', 'BirthweightCategory')

# Check frequencies of factor variables
frequency_table <- ML_5 %>%
  count(across(all_of(factor_variables)))

@@

There are still missing observations for all the remaining variables except `Date`. For now, I will keep these observations intact because I want the maximum number of observations for summary statistics of all variables. Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study. Here, I will create two datasets: one for summary statistics and one for the linear regression model. In addition, most contemporary studies exclude all birthweights less than 500 grams because they have a high chance of being stillborn. I do so the same here, even if they were marked as a live birth. I will also exclude all stillbirths or neonatal deaths as the analysis is interested in studying live births. Finally, for the linear dataset I will exclude all twin births, since most current studies restrict their regression analyses to singleton births.

```{r} 
# Create dataset for summary statistics (includes missing values)
ML_summary <- ML_5

# Creating dataset for linear regression
ML_linear <- ML_5 %>% tidyr::drop_na() %>%
  filter(Weightgrams >= 500, FetalOutcome != "morto", Gestation != "twin") # Drop observations of < 500 grams, all non-live births, and all twin births for regression dataset

# Check for no missing data in linear model
ML_linear_summary <- ML_linear %>% dplyr::summarise_all(list(~sum(is.na(.))))
ML_linear_summary
```

Now, I can see that my `ML_linear` dataset has no missing data. It has 1919 observations and 12 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.

@@

Finally, I will save the clean data both the `ML_Linear` dataset and the `ML_summary` dataset as RDS files.

```{r}
# Save ML_Linear as RDS
save_data_location <- here::here("data","processed-data","ML_linear_processed.rds")
saveRDS(ML_linear, file = save_data_location)

### 2024-06-24 products/manuscript/Manuscript_Draft.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

@@

All models based on historical data must contend with limited sample sizes and missing data. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. This means that the birthweight variable is capturing both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation. This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights.

@@

For all models in this study, the $\(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors [@leimert2020]

>>>>

This study builds on an important body of work in economic history that uses hospital records to explore trends in birth weights in the past [@costa1998, @costa2004, @goldin1989, @steckel1986, @steckel1998].

@@

All models based on historical data must contend with limited sample sizes and missing data. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. This means that the birthweight variable is capturing both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation. This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights, although other historical studies have datasets that include gestational age [@costa1998, @costa2004, @goldin1989].

@@

For all models in this study, the $\(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by genetic, environmental, and social factors [@leimert2020]. Yet, as Claudia Goldin and Robert A. Margo have shown for nineteenth-century Philadelphia, the inclusion of gestational age greatly improved the regression fit in their studies, demonstrating that gestational age could explain much of the variance of birth weight in this study [@goldin1989].
